Extended Data

### 2022-02-04 doc/manuscript/Dow_manuscript_Nature.Rmd
statement on 2.5 and 3 sd runs
Reviewer 1 suggestions

**As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014; @buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011;@finzi_carbon_2020;@stuble_plant_2021] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].**
**However, less is known about how spring temperatures affect tree stem growth, which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
**Using dendrometer band measurements from `r n_dendro_trees` trees across two forests, we show that warmer spring temperatures shifted the peak stem growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum daily growth rates, or annual growth.**
**The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual growth was far more sensitive to temperatures during the peak growing season than in the spring.**
**These findings imply that any extra CO~2~ uptake in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] does not significantly contribute to increased sequestration in long-lived woody stem biomass.**
**Rather, contradicting current projections from global C cycle models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], our empirical results imply that warming spring temperatures are unlikely to increase stem growth or strengthen the CO~2~ sink of temperate deciduous forests.**

@@

While responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been documented[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999;@finzi_carbon_2020], little is known about how the longest-lived component of fixed C in trees, the woody stem growth, is responding to warming spring temperatures. 
This uncertainty has led to  a surge in recent studies documenting the climate sensitivity of stem growth timing and its link to annual growth[@huang_photoperiod_2020; @delpierre_chilling_2019; @dorangeville_peak_2021; @etzold_number_2022; @tumajer_limitation_2022; @zweifel_why_2021]; however, the role of spring temperatures in shaping the timing of growth remains unclear.

@@

![**Figure 2 | Foliage (a,b) and stem growth (c,d) phenology at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level canopy foliage phenology from 2001-2018, obtained from the MODIS Global Vegetation Phenology product (MCD12Q2.006), where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down). Panels (c-d) show the dates at which stem growth milestones were achieved, on average, for sampled populations of ring-porous and diffuse-porous trees at SCBI (2011-2020) and Harvard Forest (1998-2003). Mean temperature was calculated for each wood-type/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red.](tables_figures/DOYtiming_allyears.png){ width=25% }

Both MODIS-derived canopy foliage phenology and dendrometer band measurements of stem growth timing generally shifted backwards as spring temperatures increased (Fig. `r DOY_timing`, `r EDF_pheno_all`). 
We found a consistent effect of temperature ($T_{max}$ or $T_{min}$) throughout the spring, but the strongest effects on stem growth timing were found using $T_{max}$ during a critical temperature window (CTW). 
CTW was identified by measuring the correlation between all combinations of weekly $T_{max}$ and $DOY_{25}$ from January 1 to mean $DOY_{25}$ for each xylem porosity-site combination (`r EDF_climwin`).
The CTW was defined as the week(s) which had the strongest correlation with $DOY_{25}$.

@@

Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem growth but have little effect on total annual growth (Figs. `r schematic_summary`- `r tree_rings`). 
The observed advance in stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig. `r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and _NEE_[@keenan_net_2014; @finzi_carbon_2020].
However, inconsistent with the concept that an earlier start to the growing season would increase annual stem growth, we demonstrate that warmer springs hasten the cessation of peak stem expansion and thereby have negligible effect on total annual stem growth for most species and locations (Fig. `r tree_rings`). 
Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020; @etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as daylength or sink limitation, which also play an important role in autumn leaf senescence[@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014].
Our tree-ring analysis (Fig. `r tree_rings`) demonstrates that the primary effect of warming temperatures on annual ring-width is not an augmentation through an earlier start to growth, but rather a reduction associated to drought stress during the peak growing season[@dorangeville_drought_2018]. This decoupling between growing season length and annual growth mirrors results from another recent study by Etzold et al[@etzold_number_2022] which found that vapor pressure deficit plays a key role in limiting stem growth *- other studies which found VPD limits growth-*.
Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021]. However, we did not find evidence that higher spring temperatures were correlated with summer drought stress in our dendrometer band analysis, implying that the effects of warm spring temperatures on stem growth elucidated here (Fig. `r schematic_summary`) were not attributable to summer drought.

Our finding that interannual variation in stem growth is more strongly linked to conditions during the peak growing season than to growing season length aligns with recent studies on annual stem growth[@etzold_number_2022; @tumajer_limitation_2022] and with parallel findings for _NEE_[@zhou_explaining_2016; @fu_maximum_2019].
However, there is also a disconnect with findings that _NEE_ is at least modestly greater in years with warm springs[@keenan_net_2014] or long growing seasons[@zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].
Warming advances spring phenology and may advance or delay autumn senescence depending on timing of warming and water availability[@zohner_ongoing_2019;@xie_predicting_2018; @crabbe_extreme_2016], with delays more common across eastern North America [@buermann_widespread_2018;@finzi_carbon_2020;@keenan_net_2014], implying that warming temperatures are lengthening the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem.
We show that the extra C fixation in years with warm springs does not substantially augment stem growth, but it remains unclear how it is allocated within the ecosystem.
There are two main possibilities, which hold contrasting implications for the response of forest C balance to rising spring temperatures.

@@

We fit a five-parameter logistic growth model[@mcmahon_general_2015] to dendrometer band data from each tree-year to define phenological dates and growth rates (Fig. `r schematic_summary`). In particular, we model the observed diameter at breast height (DBH) on a given day of the year (DOY; *i.e.*, julian days) as: 
$$ DBH = L + \frac{K - L} { 1 + 1/\theta \cdot exp[-r(DOY - DOY_{ip}) / \theta)] ^ \theta}$$ 
Here, $L$ and $K$ are lower and upper asymptotes of the model, corresponding to DBH at the beginning and end of the year, respectively. 
$DOY_{ip}$ is the day of year where the inflection point in growth rate occurs,
$r$ shapes the slope of the curve at the inflection point,
and $\theta$ is a tuning parameter controlling the slope of the curve toward the upper asymptote.
This allows an asymetric fit to the data, where the onset of growth can be estimated independent of the cessation of growth.
The DOY on which maximum growth occurs, $DOY_{g_{max}}$ (Fig. `r schematic_summary`), occurs on $DOY_{ip}$ when $\theta = 1$.
The model was fit in R v4.0 using the functions developed in the *Rdendrom* package[@mcmahon_general_2015]. These functions take the time-series of manual dendrometer band measurements and return maximum-likelihood optimized values of the above five parameters that best predict DBH for each day of year. 
An advantage of this approach is that short-term shrinkage and swelling associated with rain events [@mcmahon_general_2015; @herrmann_tree_2016] and measurement errors show up as residual variation and do not unduly influence the parameters of interest. 
Having fit the model for each band-year of data, we then modeled DBH using these optimal parameter values in our logistic growth model and extracted the intra-annual growth variables of interest (Fig. `r schematic_summary`).

After fitting the growth model, we removed tree-years with poor fits. 
Models were judged to be poorly fit if certain modeled growth characteristics fell outside of the logical range. 
Modeled fits for tree-years were removed under five conditions:
(1) single day growth rates were $\ge 2$ standard deviations away from the mean for each wood-type (SCBI = 2, Harvard Forest = 34); 
(2) $DOY_{ip}$ was $\ge 2$ standard deviations away from the mean for its xylem architecture group, year, and site (SCBI = 53, Harvard Forest = 106); 
(3) tree-years with small or negligible total growth ($\Delta DBH$ $\le 0.02$ mm; SCBI = 0, Harvard Forest = 66); 
(4) model fit predicted total yearly growth to take longer than 365 days, indicating poor model fit (SCBI = 150, Harvard Forest = 199); 
(5) models with unexplained sharp spikes in growth rate (SCBI = 0, Harvard Forest = 3); 
and (6) poorly fit models that did not meet any of the above criteria (SCBI = 2, Harvard Forest = 0).
At Harvard Forest the tag years removed through this method were proportional to the original sample size, indicating that no species or size class was disproportionately removed compared to others. 
At SCBI, a higher proportion of ring-porous trees were removed, the majority falling under condition 4.

>>>>

**As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011;@keenan_net_2014; @buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020] and commencement of net carbon dioxide (CO~2~) sequestration[@keenan_net_2014;@finzi_carbon_2020] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011;@finzi_carbon_2020;@stuble_plant_2021] and annual CO~2~ uptake[@keeling_increased_1996;@dragoni_evidence_2011;@keenan_net_2014; @crabbe_extreme_2016; @zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].**
**However, less is known about how spring temperatures affect tree stem growth[@huang_photoperiod_2020; @delpierre_chilling_2019; @rossi_critical_2008; @savage_coordination_2021], which sequesters carbon (C) in wood that has a long residence time in the ecosystem[@xue_global_2017;@russell_residence_2014].**
**Using dendrometer band measurements from `r n_dendro_trees` trees across two forests, we show that warmer spring temperatures shifted the peak stem growth of deciduous trees earlier but had no consistent effect on peak growing season length, maximum daily growth rates, or annual growth.**
**The latter finding was confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, where annual growth was far more sensitive to temperatures during the peak growing season than in the spring.**
**These findings imply that any extra CO~2~ uptake in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] does not significantly contribute to increased sequestration in long-lived woody stem biomass.**
**Rather, contradicting current projections from global C cycle models[@buermann_widespread_2018; @zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], our empirical results imply that warming spring temperatures are unlikely to increase woody productivity or strengthen the CO~2~ sink of temperate deciduous forests.**

@@

While responses of leaf phenology and seasonal _NEE_ to warming spring temperatures have been documented[@jeong_phenology_2011; @keenan_net_2014; @buermann_widespread_2018; @ibanez_forecasting_2010; @menzel_european_2006; @menzel_growing_1999;@finzi_carbon_2020], little is known about how the longest-lived component of fixed C in trees, the woody stem growth, is responding to warming spring temperatures[@huang_photoperiod_2020; @delpierre_chilling_2019; @rossi_critical_2008; @savage_coordination_2021]. 
This uncertainty has led to  a surge in recent studies documenting the climate sensitivity of stem growth timing and its link to annual growth[@dorangeville_peak_2021; @etzold_number_2022; @tumajer_limitation_2022; @zweifel_why_2021]; however, the role of spring temperatures in shaping the timing of growth remains unclear.

@@

![**Figure 2 | Foliage (a,b) and stem growth (c,d) phenology at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level canopy foliage phenology from 2001-2018, obtained from the MODIS Global Vegetation Phenology product (MCD12Q2.006), where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down). Panels (c-d) show the dates at which stem growth milestones were achieved, on average, for sampled populations of ring-porous and diffuse-porous trees at SCBI (2011-2020) and Harvard Forest (1998-2003). Mean temperature was calculated for each xylem architecture/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red.](tables_figures/DOYtiming_allyears.png){ width=25% }

Both MODIS-derived canopy foliage phenology and dendrometer band measurements of stem growth timing generally shifted backwards as spring temperatures increased (Fig. `r DOY_timing`, `r EDF_pheno_all`). 
We found a consistent effect of temperature ($T_{max}$ or $T_{min}$) throughout the spring, but the strongest effects on stem growth timing were found using $T_{max}$ during a critical temperature window (CTW). 
CTW was identified by measuring the correlation between all combinations of weekly $T_{max}$ and $DOY_{25}$ from January 1 to mean $DOY_{25}$ for each xylem architecture-site combination (`r EDF_climwin`).
The CTW was defined as the week(s) which had the strongest correlation with $DOY_{25}$.

@@

Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem growth but have little effect on total annual growth (Figs. `r schematic_summary`- `r tree_rings`). 
The observed advance in stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig. `r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021] and _NEE_[@keenan_net_2014; @finzi_carbon_2020].
However, inconsistent with the concept that an earlier start to the growing season would increase annual stem growth, we demonstrate that warmer springs hasten the cessation of peak stem expansion and thereby have negligible effect on total annual stem growth for most species and locations (Fig. `r tree_rings`). 
Our observations suggest that the cessation of rapid stem expansion, which occurs mid-summer near the time of peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020; @etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as daylength or sink limitation, which also play an important role in autumn leaf senescence[@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014].
Our tree-ring analysis (Fig. `r tree_rings`) demonstrates that the primary effect of warming temperatures on annual ring-width is not an augmentation through an earlier start to growth, but rather a reduction associated to drought stress during the peak growing season[@dorangeville_drought_2018]. This decoupling between growing season length and annual growth mirrors results from other recent studies which found that vapor pressure deficit plays a key role in limiting stem growth[@etzold_number_2022; @delpierre_wood_2016].
Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021]. However, we did not find evidence that higher spring temperatures were correlated with summer drought stress in our dendrometer band analysis, implying that the effects of warm spring temperatures on stem growth elucidated here (Fig. `r schematic_summary`) were not attributable to summer drought.

Our finding that interannual variation in stem growth is more strongly linked to conditions during the peak growing season than to growing season length aligns with recent studies on annual stem growth[@etzold_number_2022; @tumajer_limitation_2022] and with parallel findings for _NEE_[@zhou_explaining_2016; @fu_maximum_2019].
However, there is also a disconnect with findings that _NEE_ is at least modestly greater in years with warm springs[@keenan_net_2014] or long growing seasons[@zhou_explaining_2016; @fu_maximum_2019;@finzi_carbon_2020].
Warming advances spring phenology and may advance or delay autumn senescence depending on timing of warming and water availability[@zohner_ongoing_2019;@xie_predicting_2018; @crabbe_extreme_2016], with delays more common across eastern North America [@buermann_widespread_2018;@finzi_carbon_2020;@keenan_net_2014], implying that warming temperatures are lengthening the period from peak stem growth to the cessation of CO~2~ uptake by the ecosystem.
Cite delpierre 2016 around here. Delpierre et al. 2016 found that stem growth was not correlated with carbon inputs, agreeing with the results here. However, the authors 
We show that the extra C fixation in years with warm springs does not substantially augment stem growth, but it remains unclear how it is allocated within the ecosystem.
There are two main possibilities, which hold contrasting implications for the response of forest C balance to rising spring temperatures.

@@

We fit a five-parameter logistic growth model[@mcmahon_general_2015] to dendrometer band data from each tree-year to define stem growth milestones and growth rates (Fig. `r schematic_summary`). In particular, we model the observed diameter at breast height (DBH) on a given day of the year (DOY; *i.e.*, julian days) as: 
$$ DBH = L + \frac{K - L} { 1 + 1/\theta \cdot exp[-r(DOY - DOY_{ip}) / \theta)] ^ \theta}$$ 
Here, $L$ and $K$ are lower and upper asymptotes of the model, corresponding to DBH at the beginning and end of the year, respectively. 
$DOY_{ip}$ is the day of year where the inflection point in growth rate occurs,
$r$ shapes the slope of the curve at the inflection point,
and $\theta$ is a tuning parameter controlling the slope of the curve toward the upper asymptote.
This allows an asymetric fit to the data, where the onset of growth can be estimated independent of the cessation of growth.
The DOY on which maximum growth occurs, $DOY_{g_{max}}$ (Fig. `r schematic_summary`), occurs on $DOY_{ip}$ when $\theta = 1$.
The model was fit in R v4.0 using the functions developed in the *Rdendrom* package[@mcmahon_general_2015]. These functions take the time-series of manual dendrometer band measurements and return maximum-likelihood optimized values of the above five parameters that best predict DBH for each day of year. 
An advantage of this approach is that short-term shrinkage and swelling associated with rain events [@mcmahon_general_2015; @herrmann_tree_2016] and measurement errors show up as residual variation and do not unduly influence the parameters of interest. 
Having fit the model for each band-year of data, we then modeled DBH using these optimal parameter values in our logistic growth model and extracted the intra-annual growth variables of interest (Fig. `r schematic_summary`).

After fitting the growth model, we removed tree-years with poor fits. 
Models were judged to be poorly fit if certain modeled growth characteristics fell outside of the logical range. 
Modeled fits for tree-years were removed under five conditions:
(1) single day growth rates were $\ge 2$ standard deviations away from the mean for each xylem architecture (SCBI = 2, Harvard Forest = 34); 
(2) $DOY_{ip}$ was $\ge 2$ standard deviations away from the mean for its xylem architecture group, year, and site (SCBI = 53, Harvard Forest = 106); 
(3) tree-years with small or negligible total growth ($\Delta DBH$ $\le 0.02$ mm; SCBI = 0, Harvard Forest = 66); 
(4) model fit predicted total yearly growth to take longer than 365 days, indicating poor model fit (SCBI = 150, Harvard Forest = 199); 
(5) models with unexplained sharp spikes in growth rate (SCBI = 0, Harvard Forest = 3); 
and (6) poorly fit models that did not meet any of the above criteria (SCBI = 2, Harvard Forest = 0).
At Harvard Forest the tag years removed through this method were proportional to the original sample size, indicating that no species or size class was disproportionately removed compared to others. 
At SCBI, a higher proportion of ring-porous trees were removed, the majority falling under condition 4.
This process was repeated using 2.5 and 3 standard deviations and the results of analyses were not qualitatively different.
